---
title: "From Beans to Means: What Coffee Preferences Reveal About Americans"
subtitle: "Data Wrangling with 'The Great American Coffee Taste Test'"
author: "Maryam Choudhury"
date: "`r Sys.Date()`"
output:
  learnr::tutorial:
    progressive: true
    theme: flatly
    code_folding: hide
    toc: true
    toc_float: true
    code_download: true
  html_document:
    df_print: paged
runtime: shiny_prerendered
---

```{r global-setup, include=FALSE, echo=FALSE, message=FALSE}
library(tidyverse)
coffee_data <- read_csv("data/GACTT_RESULTS_ANONYMIZED.csv")
message("Global setup ran and coffee_data is available.")
```

```{r drinker-type-setup}
filtered_coffee <- coffee_data |> 
  filter(!is.na(coffee_type), !is.na(pref_score))
```

```{r pivot-coffee-ratings-setup}
filtered_coffee <- coffee_data |> 
  filter(!is.na(coffee_type), !is.na(pref_score))
```

```{r roast-pref-ex-setup}
rating_data <- coffee_data |> 
  filter(!is.na(roast_level), !is.na(pref_score)) |> 
  mutate(roast_level = factor(roast_level))
```

```{r facet-setup}
rating_data <- coffee_data |> 
  filter(!is.na(coffee_type), !is.na(pref_score), !is.na(age_group))
```

## Welcome!

This tutorial walks you through the data-wrangling process with exercises and visualizations. We will be using data from the [**Great American Coffee Taste Test**](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-05-14/readme.md), a nationwide coffee tasting initiative conducted by James Hoffmann and Cometeer in October 2023.

The **Great American Coffee Taste Test** is extremely exciting to me on the basis of the scale and efforts required to collect it's data. Over 5,000 kits were mailed to volunteers across the U.S. who blindly rated four coffees mysteriously labeled A, B, C, and D. While Hoffman's viewers may not represent the average American and their relationship with coffee (we can assume that his subscribers, for the most part, have a positive relationship with coffee as they are watching content on it for leisure), it's still still great data that could be useful for companies developing products for coffee enjoyers. In the case of an average Joe like me, it's just a good bit of fun to explore, and with it, I can offer some insight into foundational data-wrangling techniques. So, whether you are here for the data, coffee, or both, welcome!

Throughout this interactive learnr powered tutorial, we'll tidy messy survey data, reshape it into a form suitable for analysis, and use our clean and shaped data to generate visualizations. In doing so, we'll gain more insight into this question:

> **What can American coffee preferences tell us about their habits, personalities, and favorite brews?**

------------------------------------------------------------------------

## Step 0: Set Up

Before we can begin, we will need load in our necessary libraries.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)      # for data wrangling and ggplot2 - Note that the dplyr package, from which we will use multiple methods, is       included here (methods like: select() , filter() , mutate(), etc.)
library(janitor)        # for clean_names()
library(scales)         # for formatting values in plots
library(stringr)        # for string processing
library(learnr)         # makes this .rmd interactive
library(knitr)          # for formatting tables/output
library(ggthemes)       # will give us access to color-related extensions 
```

Next, we'll read our csv file into the `coffee` dataframe. Let's `View()` it. Notice the column names. Aspects like capitalization are not consistent.

```{r cleaning_setup, message=FALSE, warning=FALSE}
coffee <- read_csv("data/GACTT_RESULTS_ANONYMIZED.csv") 
View(coffee)
```

Passing `coffee` through `clean_names` will automatically convert all the variable names to lower case and put underscores in the gaps.

```{r cleaning_setup_p2, message=FALSE, warning=FALSE}
coffee <- coffee |> clean_names()  # import and clean column names
```

When we take a glimpse at the adjusted dataset, we can see that variable names have been standardized!

```{r, message=FALSE, warning=FALSE}
glimpse(coffee)
```

Note: you could have imported your data in and cleaned it all in one fell swoop (as shown below), but for illustrative purposes it's been broken up.

```{r example_code, eval=FALSE, echo=TRUE}
coffee <- read_csv("data/GACTT_RESULTS_ANONYMIZED.csv") |> 
  clean_names()
```

## Step 1: Focus Our Dataset

### What You’ll Learn:

In this section, you'll learn...

-   How to filter rows using `filter()`
-   How to select specific columns using `select()`
-   How to evaluate completeness of survey responses

We want to work only with survey respondents who actually rated coffee and gave us useful and interesting information about themselves. So let's isolate that group.

```{r filter-data}
filtered_coffee <- coffee |> 
  filter(!is.na(coffee_a_personal_preference) | 
         !is.na(coffee_b_personal_preference) |
         !is.na(coffee_c_personal_preference)) |> 
  filter(!is.na(what_is_your_age),
         !is.na(how_many_cups_of_coffee_do_you_typically_drink_per_day))
```

```{r check-before-after-filtering, echo=FALSE}
tibble(
  Before = nrow(coffee),
  After = nrow(filtered_coffee)
)
```

> ️ *We removed incomplete rows. Now we're down to `r nrow(filtered_coffee)` focused and useful responses.*

------------------------------------------------------------------------

## Step 2: Clean Up Age and Consumption

### What You’ll Learn:

-   Use `parse_number()` to convert messy numeric text fields
-   Use `mutate()` and `case_when()` to create grouped categories

```{r clean-age}
filtered_coffee <- filtered_coffee |> 
  mutate(
    age = parse_number(what_is_your_age),  # extract numeric age from mixed-format responses
    age_group = case_when(                 # group respondents into age categories
      age < 25 ~ "Under 25",
      age < 35 ~ "25-34",
      age < 45 ~ "35-44",
      age < 55 ~ "45-54",
      age >= 55 ~ "55+",
      TRUE ~ "Unknown"),
    cups_per_day = parse_number(how_many_cups_of_coffee_do_you_typically_drink_per_day)  # parse daily coffee intake
  )
```

```{r drinker-type, exercise=TRUE, setup="drinker-type-setup"}
# Try it: Use case_when() to create a "drinker_type" variable
# based on cups_per_day:
# Light (0–1 cups), Moderate (2–3 cups), Heavy (4+ cups)
filtered_coffee <- filtered_coffee |> 
  mutate(drinker_type = case_when(
    # your logic here
  ))
```

### Solution Feedback: Drinker Type Variable

```{r drinker-type-solution, echo=TRUE}
filtered_coffee <- filtered_coffee |> 
  mutate(
    cups_per_day = parse_number(how_many_cups_of_coffee_do_you_typically_drink_per_day),
    drinker_type = case_when(
      cups_per_day <= 1 ~ "Light",
      cups_per_day <= 3 ~ "Moderate",
      cups_per_day > 3 ~ "Heavy"
    )
  )
```

We took the raw numeric column 'how_many_cups_of_coffee_do_you_typically_drink_per_day' and transformed it into something more insightful: a 'drinker_type' category! In theory, this will help us meaningfully compare how different levels of coffee consumption relate to preferences or other traits down the line. Please note that the boundaries between light, moderate and heavy were a personal consideration. Others may, for instance, consider someone who drinks three cups of coffee/day to be a light coffee drinker. Thus, it is very important to clearly convey how you are manipulating your data and why!

------------------------------------------------------------------------

## Step 3: Tidy the Coffee Ratings

### What You’ll Learn:

-   How to reshape wide data into tidy long format with `pivot_longer()`
-   Use `pivot_wider()` to clean up after reshaping

If we tried to include all `coffee_*` columns, we’d get this error:

> \_ *Can't combine `coffee_a_bitterness`* <double> and `coffee_a_notes` <character>.

That’s because text fields like `notes` can't be merged with numeric scores under a shared name during the `pivot_wider()` step.

Let’s avoid that problem by including only the numeric rating columns: `bitterness`, `acidity`, and `personal_preference`.

```{r pivot-coffee-ratings, exercise = TRUE, setup="pivot-coffee-ratings-setup"}
# Try it: Select only numeric coffee score columns to avoid type errors
# Hint: Use select(matches(...)) to get only bitterness, acidity, and personal_preference

# Your code here
```

### Solution Feedback: Safe Coffee Score Pivot

```{r pivot-coffee-ratings-solution, echo=TRUE}
ratings <- filtered_coffee |> 
  select(matches("coffee_[a-d]_(bitterness|acidity|personal_preference)"), respondent_id) |>  # select only numeric coffee rating columns plus respondent ID
  pivot_longer(                                                                                  # reshape data from wide to long format
    cols = starts_with("coffee_"),                                                               # include all columns that start with "coffee_"
    names_to = c("coffee_label", "rating_type"),                                                # splits column names into two new columns: label (a-d) and rating type
    names_pattern = "coffee_([a-d])_(.*)",                                                       # use regex to extract the label and type from column names
    values_to = "rating"                                                                          # the values go into a single new column named "rating"
  ) |> 
  pivot_wider(names_from = rating_type, values_from = rating)                                   # reshape data to put rating types (bitterness, acidity, etc.) into separate columns
```

When you run the code given in the previous solution feedback section, you get the list-cols warning above. It is telling you that for at least one combination of `respondent_id`, `coffee_label`, and `rating_type`, there is more than one value. This suggests that there might be a duplicate set of responses in our datset for a single individual. How will R deal with this? Well, since `pivot_wider()` expects each combination to yield a single value, it will default to storing the multiple values in a list-column. While this isn't inherently bad it could cause unnecessary complications down the road, so we should 'wrangle' this data now. To do so, we will run the code below. It will find which combinations are duplicated, thereby allowing us to gauge if any respondent rated the same coffee more than once for the same attribute.

```{r}
ratings_check <- filtered_coffee |> 
  select(matches("coffee_[a-d]_(bitterness|acidity|personal_preference)"), respondent_id) |> 
  pivot_longer(
    cols = starts_with("coffee_"),
    names_to = c("coffee_label", "rating_type"),
    names_pattern = "coffee_([a-d])_(.*)",
    values_to = "rating"
  )

ratings_check |> 
  dplyr::summarise(n = dplyr::n(), .by = c(respondent_id, coffee_label, rating_type)) |> 
  dplyr::filter(n > 1)

```

**What does this output tell us?**

n = 2, so we know that for respondent KVG007, every coffee rating (A–D) and rating type (bitterness, acidity, personal_preference) appears twice in the dataset! Let's clean this up.

```{r}
# Our revised solution will remove duplicate ratings BEFORE pivoting wider
# This ensures each (respondent_id, coffee_label, rating_type) combo appears only once

ratings <- filtered_coffee |> 
  select(matches("coffee_[a-d]_(bitterness|acidity|personal_preference)"), respondent_id) |> 
  pivot_longer(
    cols = starts_with("coffee_"),                                # reshape wide columns like coffee_a_bitterness
    names_to = c("coffee_label", "rating_type"),                   # create two new columns from name parts
    names_pattern = "coffee_([a-d])_(.*)",                         # extract A/B/C/D and the rating type
    values_to = "rating"                                           # values go into new column 'rating'
  ) |> 
  distinct(respondent_id, coffee_label, rating_type, .keep_all = TRUE) |>  # keep only first entry if duplicates exist
  pivot_wider(
    names_from = rating_type, 
    values_from = rating
  )

```

```{r show-tidy-example, echo=FALSE}
ratings |> filter(coffee_label == "a") |> head(3)
```

As you can see from the above output, the variable `coffee_label` isn't very insightful. Someone who is not familiar with our dataset probably would not know what a,b,c, or d are representing. Since the description of the dataset, provided by James Hoffman on Kaggle, does gives us this much needed insight, let's edit our dataframe to reflect what a-d are representing, roast level!

```{r add-coffee-type}
ratings <- ratings |> 
  mutate(coffee_type = case_when(  # map A-D labels to roast types for future analysis
    coffee_label == "a" ~ "Light - Washed",
    coffee_label == "b" ~ "Medium",
    coffee_label == "c" ~ "Dark",
    coffee_label == "d" ~ "Light - Natural"
  ))

ratings |> head(3)
```

Now that we have the column `coffee_type`, one could easily get rid of `coffee_label`, depending on your future intentions for the df!

------------------------------------------------------------------------

## Step 4: Merge Demographics with Ratings

### What You’ll Learn:

-   Use `left_join()` to merge datasets
-   Understand how row structure affects join relationships
-   Learn why we need to pre-process one of the datasets before joining

Before we combine the `ratings` and `filtered_coffee` datasets, we need to think through **how many rows** each dataset has per respondent. This determines whether our join will work cleanly or will accidentally create a ton of duplicate rows (Hint: If we aren't careful, it will!)

### Why this matters

In our `ratings` dataset, each respondent appears **multiple times** (once for each coffee and rating type). That’s because we reshaped the data using `pivot_longer()`. Here's a sample:

```{r preview_ratings, echo=TRUE}
ratings |> filter(respondent_id == "KVG007")
```

At the same time, our `filtered_coffee` dataset keeps one row for every original survey response. This means some people still show up more than once, especially if they rated multiple coffees.

```{r preview_filtered, echo=TRUE}
filtered_coffee |> filter(respondent_id == "KVG007") |> 
  select(respondent_id, age_group, cups_per_day, gender, household_income)
```

Thus, if we were to `left_join()` ratings and the selected rows from `filtered_coffee` directly, R would find multiple matches on both sides and start multiplying rows in a many-to-many relationship. The result after our join would be extremely messy and produce warnings.

To avoid this, we must make sure the demographic data we’re joining only has one row per person, so that when it gets added to the ratings (which already has multiple rows per person, specifically one for each coffee), we don’t accidentally multiply things or mess up the structure. That way, the join stays tidy.

```{r join-data-fixed, message=FALSE, warning=FALSE}
# Safe join: limit demographic side to one row per person
rating_data <- ratings |> 
  left_join(
    filtered_coffee |> 
      select(respondent_id, age_group, cups_per_day, gender, household_income) |> 
      distinct(respondent_id, .keep_all = TRUE),  # keep only one demographic row per person
    by = "respondent_id"
  )
```

Let's spot-check by one respondent to make sure out join worked correctly:

```{r preview-joined, echo=TRUE}
rating_data |> 
  filter(respondent_id == "KVG007") |> 
  select(respondent_id, coffee_label, personal_preference, age_group, gender)
```

By planning ahead and limiting the right-hand side of the join to one row per person, we prevented a many-to-many join and kept our data tidy and easy to work with!

------------------------------------------------------------------------

## Visualizations

In the following steps we will further manipulate our wrangled data to generate insightful visualizations!

## Step 5: Visualize Preferences

### What You’ll Learn:

-   How to build a grouped boxplot with `ggplot2`
-   Compare coffee types by age group

We will start by generating a visualization together! Please pay attention to the comments, as they explain the process each step of the way.

```{r plot-preference-by-age}
# The age_group column is a character vector, meaning ggplot will order it alphabetically and 'Under 25' will end up last.
# So, let's explicitly order age groups so they appear in logical (chronological) order on the x-axis
rating_data <- rating_data |> 
  mutate(age_group = factor(age_group, levels = c("Under 25", "25-34", "35-44", "45-54", "55+")))

# Plotting preference scores by age group and coffee type using boxplots
rating_data |> 
  filter(!is.na(personal_preference)) |>  # Remove rows where the preference score is missing to avoid warnings
  ggplot(aes(x = age_group, y = as.numeric(personal_preference), fill = coffee_type)) + 
  geom_boxplot(
    position = position_dodge(0.8),  # Separate boxplots within each age group for readability 
    width = 0.6,                     # Slightly narrow the boxes for readability
    color = "black"                  # Add black borders to the boxes for contrast
  ) +
  labs(
    title = "Roast Preference Scores by Age Group",      # Chart title
    x = "Age Group",                               # X-axis label
    y = "Preference Rating (1–5)",                 # Y-axis label
    fill = "Coffee Type"                           # Legend title
  ) +
   scale_fill_colorblind() # Accessible theme:)

```

------------------------------------------------------------------------

## Step 6: Explore Preference by Roast Type

### What You’ll Learn:

-   Use `group_by()` and `summarize()` to get average values
-   Clean up NA values using `na.rm = TRUE`

```{r roast-pref-ex, exercise=TRUE, setup="roast-pref-ex-setup"}
# Try it: Calculate average preference score by roast type (coffee_type)
# Hint: Use group_by() and summarize()

# Your code here
```

### Solution Feedback: Grouped Preference Scores

```{r roast-pref-ex-solution, echo=TRUE}
rating_data |> 
  group_by(coffee_type) |> 
  summarize(avg_pref = mean(as.numeric(personal_preference), na.rm = TRUE))
```

We calculated the average preference score for each coffee type using `group_by()` and `summarize()`. This gives us a quick summary of which roasts performed best overall which will, in turn, help us determine which options people consistently liked more!

------------------------------------------------------------------------

## Step 7: Facet for Visual Storytelling

### What You’ll Learn:

-   Use `facet_wrap()` to split plots by a category
-   Show complex stories in a compact visual

```{r facet, exercise=TRUE, setup="facet-setup"}
# Try it: Use facet_wrap(~ coffee_type) to split a boxplot by roast style
# Hint: build off the age group plot we did before

# Your code here
```

### Solution Feedback: Faceted Boxplot

```{r facet-solution, echo=TRUE}
# Plot preference scores across age groups, separated by coffee type
rating_data |> 
  filter(!is.na(personal_preference)) |>  # Removing missing values 
  ggplot(aes(x = age_group, y = as.numeric(personal_preference), fill = coffee_type)) + 
  geom_boxplot(
    position = position_dodge(0.8),  # Keeps boxes within each facet evenly spaced for readability 
    width = 0.6,                     # Narrowing the boxes, as I noticed distinguishing them was difficult w/ the defaults
    color = "black"                  # Adding borders for contrast within each facet
  ) +
  facet_wrap(~ coffee_type) +  # Creating a separate subplot for each coffee roast type
  labs(
    title = "Preferences by Roast and Age Group",  # Plot title
    x = "Age Group",                               # X-axis label
    y = "Preference Rating (1–5)",                 # Y-axis label
    fill = "Coffee Type"                           # Legend title (optional but consistent)
  ) +
   scale_fill_colorblind()  

```

We broke the plot into separate panels for each roast using `facet_wrap()`. We now have a clear visualization with which to assess how each type was rated across age groups!

------------------------------------------------------------------------

### Conclusion: What We Learned from the Coffee Preference Data

Throughout this project, we wrangled and visualized real survey data from the Great American Coffee Taste Test dataset. In doing so, we practiced the following skills:

-   **Data Import & Cleaning**
    -   `read_csv()` to load data
    -   `clean_names()` to standardize column headers
    -   `parse_number()` to extract numbers from mixed-format responses
-   **Data Transformation**
    -   `mutate()` to create new variables like cups_per_day`and`drinker_type\`
    -   `case_when()` to classify continuous variables into categories
    -   `filter()` to exclude NA values
    -   `select()` and `distinct()` to control the structure of joined or summarized datasets
-   **Data Reshaping**
    -   `pivot_longer()` to tidy up wide-format ratings data
    -   `pivot_wider()` to re-format it for plotting (while resolving duplicates)
-   **Joining Data**
    -   `left_join()` to merge demographic variables with coffee ratings, using keys like `respondent_id`
-   **Visualization**
    -   `ggplot()` to build visual narratives
    -   `geom_boxplot()` to show preference distributions
    -   `facet_wrap()` and `position_dodge()` to highlight comparisons between groups

By cleaning and reshaping the data, we made it possible to clearly and meaningfully examine preference patterns in the dataset. The grouped and faceted boxplots gave us two complementary views of the data:

-   **The grouped boxplot** showed side-by-side roast preferences within each age group, making it easy to compare which roasts stood out most to each demographic.
-   **The faceted plot** flipped the lens, allowing us to see how each roast’s ratings varied across different age brackets.

What patterns do you see in the visualizations? Here are some of mine:

-   **Light - Natural roasts** tended to have the highest median preference scores across all age groups, particularly among younger and mid-aged drinkers.

-   **Dark and Medium roasts** saw relatively stable ratings, suggesting they may be more universally liked or expected.

-   **Preference variability** (as shown by the spread of the boxplots) seemed to be highest for Light - Natural roasts, which could indicate more polarizing flavor characteristics like fermentation or fruitiness (AN: Putting my knowledge from Hoffman's YouTube channel to work here!).

### Why This Matters

These insights could be useful for coffee brands or roasters aiming to target specific age demographics. For instance:

-   **Light - Natural roasts** might appeal most to adventurous younger drinkers but may be less consistent in satisfaction and thus a riskier business venture.

-   **Medium and Dark roasts** could be safe bets for broader audiences due to their more stable ratings across all groups.

Overall, this project demonstrates the power of tidy data practices in wrangling raw survey responses into actionable, age-targeted product insights. For someone following this tutorial, a good next step could be practicing the skills we went through by coming up with a question on the data, and then wrangling it as necessary to create a visualization able to answer that question. Would you need a visualization different than a boxplot? How would you need to adjust your data preparation for different visualization types? These are all great guiding questions to explore!

**Thank you for choosing and following this tutorial!** Let's all have a cup of coffee to celebrate!

## References

Allaire, J. J., Xie, Y., McPherson, J., Luraschi, J., Ushey, K., Atkins, A., ... & Iannone, R. (2024). *Learnr: Interactive tutorials for R*. RStudio. <https://rstudio.github.io/learnr/>

Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). *R for Data Science (2e)*. <https://r4ds.hadley.nz>

Wickham, H. (2024). *tidyverse: Easily Install and Load the Tidyverse*. <https://www.tidyverse.org/>

Rockweiler, N. (2025). *Lecture slides: STAT228 - Data Wrangling with R*. (Course materials accessed via institutional Moodle login.)

RStudio, PBC. (2024). *RMarkdown: The Definitive Guide*. <https://bookdown.org/yihui/rmarkdown/>

Posit (2024). *Using Learnr Exercises*. <https://rstudio.github.io/learnr/exercises.html>

TidyTuesday Project. (2024, May 14). *Great American Coffee Taste Test Dataset*. <https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-05-14/readme.md>
